\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-03 paper/ms.tex
Working on prediction section

To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
compute the log-determinant of $K$ and solve linear systems of the form
$K\,\bvec{b}=\bvec{y}$ for $\bvec{b}$, where $K$ is a rank-$2\,J$
semi-separable matrix.
If $K$ is positive semi-definite, these can be computed using the Cholesky
factorization of $K$:
\begin{eqnarray}
K = L\,D\,L^\T
\end{eqnarray}
where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
diagonal matrix.
We use the ansatz that $L$ has the form
\begin{eqnarray}
    L = I + \mathrm{tril} (U\,W^T)
\end{eqnarray}
where $I$ is the identity, $U$ is the matrix from above, and $W$ is an unknown
$N \times 2\,J$ matrix.
Combining this with \eq{semi-sep}, we find the equation
\begin{eqnarray}\eqlabel{cholesky}
A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
&=& \left[I + \mathrm{tril} (U\,W^T)\right]\,D\,
    {\left[I + \mathrm{tril} (U\,W^T)\right]}^\T \\
&=& D + \mathrm{tril}(U\,W^\T)\,D
    + D\,\mathrm{triu}(W\,U^\T) \nonumber\\
&& +\, \mathrm{tril}(U\,W^\T)\,D\,\mathrm{triu}(W\,U^\T) \nonumber
\end{eqnarray}
that we can solve for $W$.
Setting each element on the left side of \eq{cholesky} equal to the
corresponding element on the right side, we derive the following recursion
relations
\begin{eqnarray}
S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,W_{n-1,j}\,W_{n-1,k} \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
    \nonumber\\
W_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
    \eqlabel{algorithm}
\end{eqnarray}
where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
$S_{1,j,k}=0$.
The computational cost of one iteration of the update in \eq{algorithm} is
$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
observation~--~so the total cost of this factorization scales as
$\mathcal{O}(N\,J^2)$.

@@

The inverse of $K$ can be applied using the relationship
\begin{eqnarray}
\bvec{b} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
\end{eqnarray}
First, we solve for $\bvec{z} = L^{-1}\,\bvec{y}$ using the following forward
substitution algorithm
\begin{eqnarray}
    f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} +
    \tilde{W}_{n-1,j}\,z_{n-1}\right] \nonumber\\
    z_n &=& y_n - \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
\end{eqnarray}
where $f_{0,j} = 0$ for all $j$.
Using this result and the following backward substitution algorithm, we can
compute $\bvec{b}$
\begin{eqnarray}
g_{n,j} &=& \phi_{n+1,j}\,\left[g_{n+1,j} +
    \tilde{U}_{n+1,j}\,b_{n+1}\right] \nonumber\\
b_n &=& \frac{z_n}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{W}_{n,j}\,g_{n,j}
\end{eqnarray}
where $g_{N+1,j} = 0$ for all $j$.
The total cost of the forward and backward substitution passes scales as
$\mathcal{O}(N\,J)$.

@@

Simulation of a Gaussian process for large $N$ requires multiplying an
$N$-dimensional vector of Gaussian normal deviates, $\bf{q}$, by the square root
of the covariance matrix:
\begin{equation}
{\bf y} = K_\bvec{\alpha}^{1/2} {\bf q},
\end{equation}
where $q_i \sim N(0,1)$.  Note that $K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{1/2}\right)^T$, which
comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the exponent of
the likelihood with $\mu_\bvec{\theta} =0$ is given by
\begin{eqnarray}
Q &=&-\frac{1}{2}{\bf y}^T K_\bvec{\alpha}^{-1} {\bf y}\\
&=& -\frac{1}{2}{\bf y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} {\bf y}\\
&=& -\frac{1}{2}{\bf q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} {\bf q}\\
&=& -\frac{1}{2}{\bf q}^T {\bf q}.
\end{eqnarray}
The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $Q \sim \chi_N^2$.

To generate a draw from a Gaussian process with a specified \celerite\ model requires carrying out the
low-rank Cholesky decomposition first, generating the vector ${\bf q}$ of random normal deviates, and then multiplying
\begin{equation}
{\bf y} = L D^{1/2} {\bf q}.
\end{equation}
This may be carried out in $\mathcal{O}(JN)$ operations with
\begin{eqnarray}
f_{n,j} &=& \phi_{n-1,j} \left(f_{n-1,j} + \tilde{W}_{n-1,j} D_{n-1,n-1}^{1/2} q_{n-1}\right)\\
y_n &=& D_{n,n}^{1/2} q_n + \sum_{k=1}^{2J} \tilde{U}_{n,k} f_{n,k}
\end{eqnarray}
for $n = 1$ to $N$ and $j=1$ to $2J$, with $f_{0,j} = 0$ for all $j$ and $q_0 = 0$.
\agoltodo{Check that these equations are in fact correct.}
We expect that this simulation algorithm will be useful for test-driving Gaussian
process inference algorithms applied to large datasets; even simulation of
Gaussian processes becomes prohibitive for large $N$ if standard Cholesky decomposition
and multiplication are used.

@@

Multiplication of the full covariance matrix times a vector can be accomplished in $\mathcal{O}(JN)$
operations with a stable recursive algorithm using the semi-separable form:
\begin{equation}
{\bf y} =  \left(A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T)\right) {\bf z}.
\end{equation}
The algorithm for multiplication is given by
\begin{eqnarray}
f^+_{n,j} &=& \phi_{n+1,j} \left(f^+_{n+1,j} + \sum_{j=1}^J \tilde U_{n+1,j} z_{n+1}\right)\\
f^-_{n,j} &=& \phi_{n,j} \left(f^-_{n-1,j} + \sum_{j=1}^J \tilde V_{n-1,j} z_{n-1}\right)\\
y_n &=& A_{n,n} z_n + \sum_{j=1}^J \left[ \tilde V_{n,j} f^+_{n,j} + \tilde U_{n,j} f^-_{n,j}\right],
\end{eqnarray}
where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$ for all $j=1,...,J$.  The recursion requires
two sweeps, a sweep downward from $n=N-1$ to $n=1$ to compute $f^+_{n,j}$ and a sweep upwards
from $n=2$ to $n=N$ to compute $f^-_{n,j}$.  \dfmtodo{I'm getting confused by zero-based
indexing in your C code...  this may be worth scrutinizing to see if I screwed up.}

Multiplication of the full covariance matrix is useful, for example, for isolating
a particular component of a kernel, which amounts to a customized filter of a dataset.
For example, a white-noise filter is given by:
\begin{eqnarray}
\bar f &=& K({\bf t},{\bf t}) \left[K({\bf t},{\bf t})+\sigma^2{\bf I}\right]^{-1} {\bf y},\\
&=& K({\bf t},{\bf t}) {\bf b}
\end{eqnarray}
where ${\bf y}$ is the vector of measurements and ${\bf b}$ may be computed from the Cholesky
decomposition as described above.  The second step multiplication of the white-noise-free
kernel times ${\bf b}$ may be accomplished in $\mathcal{O}(JN)$ steps with this multiplication algorithm.
Note that for this particular example it is more efficient to compute $\bar f = {\bf y} -\sigma^2 {\bf I} {\bf b}$,
but in general the filtering kernel will not differ from the full kernel by a diagonal matrix.

@@

For predicting or interpolating missing or future data points (and their variance) at times ${\bf t_*} = \left\{t_{*,1},t_{*,2},...,t_{*,M}\right\}$, we need to apply the covariance between the measured and predicted times.  The mean of the predicted values is (equation 2.19 from Rasmussen \& Williams):
\begin{eqnarray}
\bar f_* &=& K({\bf t_*},{\bf t}) \left[K({\bf t},{\bf t})+\sigma^2{\bf I}\right]^{-1} {\bf y},\\
&=& K({\bf t_*},{\bf t}) {\bf b}
\end{eqnarray}
where ${\bf y}$ is the vector of measurements and $\sigma$ is the measurement uncertainty, and
${\bf b} = \left[K({\bf t},{\bf t})+\sigma^2{\bf I}\right]^{-1} {\bf y}$ can be computed from
the Cholesky decomposition of the full covariance matrix in ${\cal O}(J^2N)$.

Since the covariance matrix may be written as the sum of exponentials, we can write:
\begin{eqnarray}
\bar f_{*,i} &=& K(t_{*,i},{\bf t}) {\bf b} \\
&=& \sum_n \sum_p \alpha_p e^{-\beta_p \vert t_{*,i}-t_n\vert} b_j.
\end{eqnarray}
\agoltodo{Convert this to real, \celerite\ notation .}
Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t_{*,i}$ and
$\{t_{n0+1},...,t_N\} \ge t_{*,i}$, gives
\begin{equation}
\bar f_{*,i} = \sum_{n=1}^{n_0} \sum_p \alpha_p e^{-\beta_p( t_{*,i}-t_n)} b_n + \sum_{n=n_0+1}^N \sum_p \alpha_p e^{-\beta_p(t_n-t_{*,i})} b_n.
\end{equation}
Each of these double-sums may have the sums interchanged, giving:
\begin{eqnarray}
\bar f_{*,i} &=& \sum_p e^{-\beta_p ( t_{*,i}-t_{*,i-1})} \sum_{n=1}^{n_0} \alpha_p e^{-\beta_p ( t_{*,i-1}-t_n) } b_n \\
&+&
\sum_p e^{-\beta_p ( t_{*,i+1}-t_{*,i})} \sum_{n=n+0+1}^{N} \alpha_p e^{-\beta_p ( t_n-t_{*,i+1}) } b_n
\end{eqnarray}
We can compute this recursively.  Defining $Q^-_{i,n_0,p} = \sum_{j=1}^{n_0} \alpha_p e^{-\beta_p (t_{*,i}-t_j)} b_j$ and defining $Q^+_{i,n_0,p} = \sum_{j=n_0}^N \alpha_p e^{-\beta_p (t_{*,i}-t_j)} b_j$ gives:
\begin{equation}
\bar f_{*,i} = \sum_p e^{-\beta_p (t_{*,i}-t_{*,i-1})} Q^-_{i-1,n_0,p} + \sum_p e^{-\beta_p (t_{*,i+1}-t_{*,i})} Q^+_{i+1,n_0,p}.
\end{equation}
Once $Q^-_{1,p}$ and $Q^+_{N,p}$ are obtained in $\mathcal{O}(JN)$ operations, then the subsequent $Q^-_{i,p}$ and $Q^+_{i,p}$ values can be obtained in $\approx \mathcal{O}(JM)$ operations
\begin{eqnarray}
Q^-_{i,p} &=& e^{-\beta_p (t_{*,i}-t_{*,i-1})} Q^-_{i,p},\\
Q^+_{i,p} &=& e^{-\beta_p (t_{*,i+1}-t_{*,i})} Q^+_{i,p},
\end{eqnarray}
for a total scaling of $\approx \mathcal{O}(J^2N+J(N+M))$.

The variance evaluated at future predicted times ${\bf t_*} > {\bf t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
\begin{eqnarray}
\mathrm{var}(f_{*,i+1}) &=& K(t_{*,i+1},t_{*,i+1}) - K(t_{*,i+1},{\bf t}^T)K({\bf t},{\bf t})^{-1} K({\bf t},t_{*,i+1})\\
&=& K(t_{*,i+1},t_{*,i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} \\
&\times& e^{-\beta_p(t_{*,i}-{\bf t}^T)}K({\bf t},{\bf t})^{-1} e^{-\beta_q(t_{*,i}-{\bf t})}\\
&=& K(t_{*,i+1},t_{*,i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} P_{i,p,q}\\
\end{eqnarray}
for $1 \le p,q \le J$,
where
\begin{equation}
P_{i,p,q} = e^{-\beta_p(t_{*,i}-{\bf t}^T)}
K({\bf t},{\bf t})^{-1} e^{-\beta_q(t_{*,i}-{\bf t})}.
\end{equation}
For $i=1$, $P_{1,p,q}$ can be obtained in $\mathcal{O}(J^2N)$ operations
with the Cholesky decomposition.
Subsequent $P_{i+1,p,q}$ may be obtained recursively as
\begin{equation}
P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t_{*,i+1}- t_{*,i})} P_{i,p,q},
\end{equation}
in $\mathcal{O}(J^2M)$ operations, for a total of $\approx \mathcal{O}(J^2(N+M))$ operations.
Note that this variance formula {\it only} works for prediction; we have
yet to derive a formula for which ${\bf t_*}$ and ${\bf t}$ are
interspersed.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/astero/astero-corner.pdf}
\caption{The probabilistic constraints on $\nu_\mathrm{max}$ and $\Delta \nu$
    from the inference shown in \Figure{astero} compared to the published
    value (error bar) based on several years of \kepler\ observations
    \citep{Pinsonneault:2014}.
    \figurelabel{astero-corner}}
\end{center}
\end{figure}

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/astero/astero-comp.pdf}
\caption{A comparison between the Lomb-Scargle estimator of the PSD and the
posterior inference of the PSD as a mixture of stochastically-driven simple
harmonic oscillators.
(top) The periodogram of the \kepler\ light curve for KIC~11615890 computed
    on the full four year baseline of the mission. The orange line shows a
    smoothed periodogram and the blue line indicates the level of the
    measurement uncertainties.
(middle) The same periodogram computed using about a month of data.
(bottom) The power spectrum inferred using the mixture of SHOs model described
    in the text and only one month of \kepler\ data.
    The black line shows the median of posterior PSD and the gray contours
    show the 68\% credible region.
    \figurelabel{astero}}
\end{center}
\end{figure}

@@

In this example, we inject the signal of a simulated exoplanet transit into a
real \kepler\ light curve and then demonstrate that we can recover the true
physical parameters of the exoplanet while modeling the stellar variability
using \celerite.
This example is different from all the previous examples~--~both simulated and
real~--~because, in this case, we are uninterested in the parameters of the
covariance model.
Instead, we're interested in inferring constraints on the parameters of the
mean model ($\mu_\bvec{\theta}(\bvec{x})$ from \sect{gps}), the model of the
signal caused by the transit of the planet.
We say ``mean model'' here to unambiguously distinguish it from the
``covariance model'', but this is what would generally be called the ``model''
in other analyses.
In this case, we have a light curve of a rotating star with a planet
transiting and we would like to fit for the physical parameters of the planet
while taking our uncertainty about the star's variability into account.
GP modeling has been used for this purpose throughout the exoplanet literature
\citep[for example][]{Dawson:2014, Barclay:2015, Evans:2015,
Foreman-Mackey:2016b, Grunblatt:2016}.

>>>>

To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
compute the log-determinant of $K$ and solve linear systems of the form
$K\,\bvec{z}=\bvec{y}$ for $\bvec{z}$, where $K$ is a rank-$2\,J$
semi-separable matrix.
If $K$ is positive semi-definite, these can be computed using the Cholesky
factorization of $K$:
\begin{eqnarray}
K = L\,D\,L^\T
\end{eqnarray}
where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
diagonal matrix.
We use the ansatz that $L$ has the form
\begin{eqnarray}
    L = I + \mathrm{tril} (U\,W^T)
\end{eqnarray}
where $I$ is the identity, $U$ is the matrix from above, and $W$ is an unknown
$N \times 2\,J$ matrix.
Combining this with \eq{semi-sep}, we find the equation
\begin{eqnarray}\eqlabel{cholesky}
A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
&=& \left[I + \mathrm{tril} (U\,W^T)\right]\,D\,
    {\left[I + \mathrm{tril} (U\,W^T)\right]}^\T \\
&=& D + \mathrm{tril}(U\,W^\T)\,D
    + D\,\mathrm{triu}(W\,U^\T) \nonumber\\
&& +\, \mathrm{tril}(U\,W^\T)\,D\,\mathrm{triu}(W\,U^\T) \nonumber
\end{eqnarray}
that we can solve for $W$.
Setting each element on the left side of \eq{cholesky} equal to the
corresponding element on the right side, we derive the following recursion
relations
\begin{eqnarray}
S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,W_{n-1,j}\,W_{n-1,k} \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
    \nonumber\\
W_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
    \eqlabel{algorithm}
\end{eqnarray}
where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
$S_{1,j,k}=0$.
The computational cost of one iteration of the update in \eq{algorithm} is
$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
observation~--~so the total cost of this factorization scales as
$\mathcal{O}(N\,J^2)$.

@@

The inverse of $K$ can be applied using the relationship
\begin{eqnarray}
\bvec{z} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
\end{eqnarray}
First, we solve for $\bvec{z}^\prime = L^{-1}\,\bvec{y}$ using the following forward
substitution algorithm
\begin{eqnarray}
    f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} +
    \tilde{W}_{n-1,j}\,z_{n-1}^\prime\right] \nonumber\\
    z_n^\prime &=& y_n - \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
\end{eqnarray}
where $f_{0,j} = 0$ for all $j$.
Using this result and the following backward substitution algorithm, we can
compute $\bvec{z}$
\begin{eqnarray}
g_{n,j} &=& \phi_{n+1,j}\,\left[g_{n+1,j} +
    \tilde{U}_{n+1,j}\,z_{n+1}\right] \nonumber\\
z_n &=& \frac{z_n^\prime}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{W}_{n,j}\,g_{n,j}
\end{eqnarray}
where $g_{N+1,j} = 0$ for all $j$.
The total cost of the forward and backward substitution passes scales as
$\mathcal{O}(N\,J)$.

@@

Simulation of a Gaussian process for large $N$ requires multiplying an
$N$-dimensional vector of Gaussian normal deviates, $\bvec{q}$, by the square root
of the covariance matrix:
\begin{equation}
\bvec{y} = K_\bvec{\alpha}^{1/2} \bvec{q},
\end{equation}
where $q_i \sim N(0,1)$.  Note that $K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{1/2}\right)^T$, which
comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the exponent of
the likelihood with $\mu_\bvec{\theta} =0$ is given by
\begin{eqnarray}
\mathcal{Q} &=&\bvec{y}^T K_\bvec{\alpha}^{-1} \bvec{y}\\
&=& \bvec{y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} \bvec{y}\\
&=& \bvec{q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} \bvec{q}\\
&=& \bvec{q}^T \bvec{q}.
\end{eqnarray}
The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $Q \sim \chi_N^2$.

To generate a draw from a Gaussian process with a specified \celerite\ model requires carrying out the
low-rank Cholesky decomposition first, generating the vector $\bvec{q}$ of random normal deviates, and then multiplying
\begin{equation}
\bvec{y} = L D^{1/2} \bvec{q}.
\end{equation}
This may be carried out in $\mathcal{O}(JN)$ operations with
\begin{eqnarray}
f_{n,j} &=& \phi_{n-1,j} \left(f_{n-1,j} + \tilde{W}_{n-1,j} D_{n-1,n-1}^{1/2} q_{n-1}\right)\\
y_n &=& D_{n,n}^{1/2} q_n + \sum_{k=1}^{2J} \tilde{U}_{n,k} f_{n,k}
\end{eqnarray}
for $n = 1$ to $N$ and $j=1$ to $2J$, with $f_{0,j} = 0$ for all $j$ and $q_0 = 0$.
\agoltodo{Check that these equations are in fact correct.}
We expect that this simulation algorithm will be useful for test-driving Gaussian
process inference algorithms applied to large datasets; even simulation of
Gaussian processes becomes prohibitive for large $N$ if standard Cholesky decomposition
and multiplication are used.

@@

Multiplication of the full covariance matrix times a vector can be accomplished in $\mathcal{O}(JN)$
operations with a stable recursive algorithm using the semi-separable form:
\begin{equation}
\bvec{y} =  \left(A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T)\right) \bvec{z}.
\end{equation}
The algorithm for multiplication is given by
\begin{eqnarray}
f^+_{n,j} &=& \phi_{n+1,j} \left(f^+_{n+1,j} + \sum_{j=1}^J \tilde U_{n+1,j} z_{n+1}\right)\\
f^-_{n,j} &=& \phi_{n,j} \left(f^-_{n-1,j} + \sum_{j=1}^J \tilde V_{n-1,j} z_{n-1}\right)\\
y_n &=& A_{n,n} z_n + \sum_{j=1}^J \left[ \tilde V_{n,j} f^+_{n,j} + \tilde U_{n,j} f^-_{n,j}\right],
\end{eqnarray}
where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$ for all $j=1,...,J$.  The recursion requires
two sweeps, a sweep downward from $n=N-1$ to $n=1$ to compute $f^+_{n,j}$ and a sweep upwards
from $n=2$ to $n=N$ to compute $f^-_{n,j}$.  \dfmtodo{I'm getting confused by zero-based
indexing in your C code...  this may be worth scrutinizing to see if I screwed up.}

Multiplication of the full covariance matrix is useful, for example, for isolating
a particular component of a kernel, which amounts to a customized filter of a dataset.
For example, a white-noise filter is given by:
\begin{eqnarray}
\bar f &=& K(\bvec{t},\bvec{t}) \left[K(\bvec{t},\bvec{t})+\sigma^2I\right]^{-1} \bvec{y},\\
&=& K(\bvec{t},\bvec{t}) \bvec{z}
\end{eqnarray}
where $\bvec{y}$ is the vector of measurements and $\bvec{z}$ may be computed from the Cholesky
decomposition as described above.  The second step multiplication of the white-noise-free
kernel times $\bvec{z}$ may be accomplished in $\mathcal{O}(JN)$ steps with this multiplication algorithm.
Note that for this particular example it is more efficient to compute $\bar f = \bvec{y} -\sigma^2 I \bvec{z}$,
but in general the filtering kernel will not differ from the full kernel by a diagonal matrix.

@@

For predicting or interpolating missing or future data points (and their variance) at times $\bvec{t_*} = \left\{t_{*,1},t_{*,2},...,t_{*,M}\right\}$, we need to apply the covariance between the measured and predicted times.  The mean of the predicted values is (equation 2.19 from Rasmussen \& Williams):
\begin{eqnarray}
\bar f_* &=& K(\bvec{t_*},\bvec{t}) \left[K(\bvec{t},\bvec{t})+\sigma^2 I\right]^{-1} \bvec{y},\\
&=& K(\bvec{t_*},\bvec{t}) \bvec{z}
\end{eqnarray}
where $\bvec{y}$ is the vector of measurements and $\sigma$ is the measurement uncertainty, and
$\bvec{z} = \left[K(\bvec{t},\bvec{t})+\sigma^2 I\right]^{-1} \bvec{y}$ can be computed from
the Cholesky decomposition of the full covariance matrix in ${\cal O}(J^2N)$.

Since the covariance matrix may be written as the sum of exponentials, we can write:
\begin{eqnarray}
\bar f_{*,i} &=& K(t_{*,i},\bvec{t}) \bvec{z} \\
&=& \sum_n \sum_{j=1}^J e^{-c_j \vert t_{*,i}-t_n\vert}\left(a_j \cos{(d_j \vert t_{*,i}-t_n\vert)} +b_j \sin{(d_j \vert t_{*,i}-t_n\vert)}\right) z_j.
\end{eqnarray}
Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t_{*,i}$ and
$\{t_{n0+1},...,t_N\} \ge t_{*,i}$, gives
\begin{eqnarray}
\bar f_{*,i} &=& \sum_{j=1}^J\sum_{n=1}^{n_0} e^{-c_j(t_{*,i}-t_n)} \left[a_j \cos{(d_j(t_{*,i}-t_n))}+b_j\sin{(d_j(t_{*,i}-t_n))}\right] z_n \\
&+& \sum_{j=1}^J\sum_{n=n_0+1}^N  e^{-c_j(t_n-t_{*,i})} \left[a_j \cos{(d_j(t_n-t_{*,i}))}+b_j\sin{(d_j(t_n-t_{*,i}))}\right] z_n.
\end{eqnarray}
%Each of these double-sums may have the sums interchanged, giving:
%\begin{eqnarray}
%\bar f_{*,i} &=& Re\left[\sum_j e^{-(c_j+id_j)( t_{*,i}-t_{*,i-1})} \sum_{n=1}^{n_0} (a_j+ib_j) e^{-(c_j+id_j)( t_{*,i-1}-t_n) } z_n\right. \\
%&+&\left.
%\sum_j e^{-(c_j+id_j)( t_{*,i+1}-t_{*,i})} \sum_{n=n_0+1}^{N} (a_j+ib_j) e^{-(c_j+id_j)( t_n-t_{*,i+1}) } z_n\right]
%\end{eqnarray}
We can compute this recursively.  Defining
\begin{eqnarray}
Q^-_{i,n_0,2j-1} &=& \sum_{n=1}^{n_0} a_j e^{-c_j(t_{*,i}-t_n)} \cos{(d_j(t_{*,i}-t_n))} z_n,\\
Q^-_{i,n_0,2j} &=& \sum_{n=1}^{n_0} b_j e^{-c_j(t_{*,i}-t_n)} \sin{(d_j(t_{*,i}-t_n))} z_n,\\
Q^+_{i,n_0,2j-1} &=& \sum_{n=n_0+1}^N a_j e^{-c_j(t_n-t_{*,i})}\cos{(d_j(t_n-t_{*,i}))} z_n,\\
Q^+_{i,n_0,2j} &=& \sum_{n=n_0+1}^N b_j e^{-c_j(t_n-t_{*,i})} \sin{(d_j(t_n-t_{*,i}))} z_n,
\end{eqnarray} gives:
\agoltodo{Convert the following to real, \celerite\ notation .}
\begin{equation}
\bar f_{*,i} = \sum_p e^{-\beta_p (t_{*,i}-t_{*,i-1})} Q^-_{i-1,n_0,p} + \sum_p e^{-\beta_p (t_{*,i+1}-t_{*,i})} Q^+_{i+1,n_0,p}.
\end{equation}
Once $Q^-_{1,p}$ and $Q^+_{N,p}$ are obtained in $\mathcal{O}(JN)$ operations, then the subsequent $Q^-_{i,p}$ and $Q^+_{i,p}$ values can be obtained in $\approx \mathcal{O}(JM)$ operations
\begin{eqnarray}
Q^-_{i,p} &=& e^{-\beta_p (t_{*,i}-t_{*,i-1})} Q^-_{i,p},\\
Q^+_{i,p} &=& e^{-\beta_p (t_{*,i+1}-t_{*,i})} Q^+_{i,p},
\end{eqnarray}
for a total scaling of $\approx \mathcal{O}(J^2N+J(N+M))$.

The variance evaluated at future predicted times $\bvec{t_*} > \bvec{t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
\begin{eqnarray}
\mathrm{var}(f_{*,i+1}) &=& K(t_{*,i+1},t_{*,i+1}) - K(t_{*,i+1},\bvec{t}^T)K(\bvec{t},\bvec{t})^{-1} K(\bvec{t},t_{*,i+1})\\
&=& K(t_{*,i+1},t_{*,i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} \\
&\times& e^{-\beta_p(t_{*,i}-\bvec{t}^T)}K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t_{*,i}-\bvec{t})}\\
&=& K(t_{*,i+1},t_{*,i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} P_{i,p,q}\\
\end{eqnarray}
for $1 \le p,q \le J$,
where
\begin{equation}
P_{i,p,q} = e^{-\beta_p(t_{*,i}-\bvec{t}^T)}
K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t_{*,i}-\bvec{t})}.
\end{equation}
For $i=1$, $P_{1,p,q}$ can be obtained in $\mathcal{O}(J^2N)$ operations
with the Cholesky decomposition.
Subsequent $P_{i+1,p,q}$ may be obtained recursively as
\begin{equation}
P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t_{*,i+1}- t_{*,i})} P_{i,p,q},
\end{equation}
in $\mathcal{O}(J^2M)$ operations, for a total of $\approx \mathcal{O}(J^2(N+M))$ operations.
Note that this variance formula {\it only} works for prediction; we have
yet to derive a formula for which $\bvec{t_*}$ and $\bvec{t}$ are
interspersed.

@@

\begin{figure}[!p]
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/astero/astero-corner.pdf}
\caption{The probabilistic constraints on $\nu_\mathrm{max}$ and $\Delta \nu$
    from the inference shown in \Figure{astero} compared to the published
    value (error bar) based on several years of \kepler\ observations
    \citep{Pinsonneault:2014}.
    The two-dimensional contours show the 0.5-, 1-, 1.5, and 2-sigma credible
    regions in the marginalized planes and the histograms along the diagonal
    show the marginalized posterior for each parameter.
    \figurelabel{astero-corner}}
\end{center}
\end{figure}

\begin{figure}[!p]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/astero/astero-comp.pdf}
\caption{A comparison between the Lomb-Scargle estimator of the PSD and the
posterior inference of the PSD as a mixture of stochastically-driven simple
harmonic oscillators.
(top) The periodogram of the \kepler\ light curve for KIC~11615890 computed
    on the full four year baseline of the mission. The orange line shows a
    smoothed periodogram and the blue line indicates the level of the
    measurement uncertainties.
(middle) The same periodogram computed using about a month of data.
(bottom) The power spectrum inferred using the mixture of SHOs model described
    in the text and only one month of \kepler\ data.
    The black line shows the median of posterior PSD and the gray contours
    show the 68\% credible region.
    \figurelabel{astero}}
\end{center}
\end{figure}

@@

\response{
This example is different from all the previous examples~--~both simulated and
real~--~because, in this case, do not set the deterministic
mean function $\mu_\bvec{\theta}$ to zero.
Instead, we make inferences about $\mu_\bvec{\theta}$ because it is a
physically significant function and the parameters are fundamental properties
of the system.
This is an example using real data~--~because we use the light curve from
\sect{rotation}~--~but we multiply these data by a simulated transiting
exoplanet model with known parameters.
This allows us to show that we can recover the true parameters of the planet
even when the transit signal is superimposed on the real variability of a
star.
}
GP modeling has been used extensively for this purpose throughout the
exoplanet literature \citep[for example][]{Dawson:2014, Barclay:2015,
Evans:2015, Foreman-Mackey:2016b, Grunblatt:2016}.

In \eq{gp-likelihood} the physical parameters of the exoplanet are called
$\bvec{\theta}$ and, in this example, the mean function $\mu_\bvec{\theta}(t)$
is a limb-darkened transit light curve \citep{Mandel:2002} and the parameters
$\bvec{\theta}$ are the orbital period $P_\mathrm{orb}$, the transit duration
$T$, the phase or epoch $t_0$, the impact parameter $b$, the radius of the
planet in units of the stellar radius $R_P/R_\star$, and two parameters
describing the limb-darkening profile of the star
\citep{Claret:2011,Kipping:2013}.
As in \sect{rotation}, we model the stellar variability using a GP model
with a kernel given by \eq{rot-kernel} and fit for the parameters of the
exoplanet $\bvec{\theta}$ and the stellar variability $\bvec{\alpha}$
simultaneously.
\response{The full set of parameters $\bvec{\alpha}$ and $\bvec{\theta}$ are
listed in Table~\ref{tab:example-5-params} along with their priors and the
true values for $\bvec{\theta}$.}
